---
title: "CMI-PB batchQC Correction"
output: html_notebook
author: Joaquin Reyna (jreyna@lji.org)
---

## Import libraries
```{r, import-libraries}
library(BatchQC)
library(sva)
library(reshape2)
library(ggplot2)

# paste0(as.vector(rnaseq_baseline_mat_imputed_20[,'subject_id']), rep('/', nrow(rnaseq_baseline_mat_imputed_20)),as.vector(rnaseq_baseline_mat_imputed_20[,'planned_day_relative_to_boost']))
```

## Set environment values
```{r, set-env}
# set year
if (exists("params") == TRUE) {
    assay <- as.integer(params$assay)
} else {
    assay <- "rnaseq"
}
message(paste("Assay chosen:", assay))

# Set path to the harmonized directory
imputed_dir = "../../../results/main/cmi_pb_datasets/processed/all_versus_all/imputation/test/"

# Create output directory
outdir <- "../../../results/main/cmi_pb_datasets/processed/all_versus_all/batchqc/"
dir.create(outdir, recursive = TRUE, showWarnings = TRUE)
```

## Load imputed data
```{r}
# imputed_dir is path to local drive where data is saved
#setwd(imputed_dir) 

load_data <- function(fn) {
    df <- read.table(rnaseq_fn, header = 1)
    idx_names <- paste0(as.vector(df[,'subject_id']),
                        rep('/', nrow(df)),
                        as.vector(df[,'planned_day_relative_to_boost']))
    row.names(df) <- idx_names
    df <- subset(df, select=-c(subject_id, planned_day_relative_to_boost))
    return(df)
}

# import 2020 imputed datasets
tmpl_fn = file.path(imputed_dir, '%s.%d.tsv')

rnaseq_fn = sprintf(tmpl_fn, "rnaseq", 2020)
rnaseq_baseline_mat_imputed_20 <- load_data(rnaseq_fn)

cytof_fn = sprintf(tmpl_fn, "live_cell_percentages", 2020)
cytof_baseline_mat_imputed_20 <- load_data(cytof_fn)

olink_fn = sprintf(tmpl_fn, "olink", 2020)
olink_baseline_mat_imputed_20 <- load_data(olink_fn)

abtiters_fn = sprintf(tmpl_fn, "ab_titer", 2020)
abtiters_baseline_mat_imputed_20 <- load_data(abtiters_fn)

# import 2021 imputed datasets
rnaseq_fn = sprintf(tmpl_fn, "rnaseq", 2021)
rnaseq_baseline_mat_imputed_21 <- load_data(rnaseq_fn)

cytof_fn = sprintf(tmpl_fn, "live_cell_percentages", 2021)
cytof_baseline_mat_imputed_21 <- load_data(cytof_fn)

olink_fn = sprintf(tmpl_fn, "olink", 2021)
olink_baseline_mat_imputed_21 <- load_data(olink_fn)

abtiters_fn = sprintf(tmpl_fn, "ab_titer", 2021)
abtiters_baseline_mat_imputed_21 <- load_data(abtiters_fn)

# import 2022 imputed datasets
#rnaseq_fn = sprintf(tmpl_fn, "rnaseq", 2022)
#rnaseq_baseline_mat_imputed_22 <- read.csv(rnaseq_fn, row.names=1)
#
#cytof_fn = sprintf(tmpl_fn, "live_cell_percentages", 2022)
#cytof_baseline_mat_imputed_22 <- read.csv(cytof_fn, row.names=1)
#
#cytof_fn = sprintf(tmpl_fn, "olink", 2022)
#olink_baseline_mat_imputed_22 <- read.csv(olink_fn, row.names=1)
#
#cytof_fn = sprintf(tmpl_fn, "ab_titer", 2022)
#abtiters_baseline_mat_imputed_22 <- read.csv(abtiters_fn, row.names=1)
```

`BatchQC` is a package that allows assessment of data for batch effects.  
For more information, see [https://academic.oup.com/bioinformatics/article/32/24/3836/2525651]
`ComBat` is a method to correct for batch effects, see[https://rdrr.io/bioc/sva/man/ComBat.html].  
There are many different options for Combat, depending on the distribution of your data.  
To help assess whether and how to use ComBat, we've created a helpful flow chart, [https://raw.githubusercontent.com/CMI-PB/2020_2021_data_merge/main/figures/BatchCorrection_Flowchart.png]

## Batch QC for cell frequency
### Prepare data
```{r}
# Data merge
cytof_all <- cbind(t(cytof_baseline_mat_imputed_20),
                   t(cytof_baseline_mat_imputed_21))
batch  <- c (rep('2020', nrow(cytof_baseline_mat_imputed_20)),
             rep('2021', nrow(cytof_baseline_mat_imputed_21)))
```

```{r}
cytof_20 <- t(cytof_baseline_mat_imputed_20)
cytof_21 <- t(cytof_baseline_mat_imputed_21)
mcytof_20 <- melt(cytof_20)
mcytof_21 <- melt(cytof_21)
mcytof_20$year <- "2020"
mcytof_21$year <- "2021"
mcytof <- rbind(mcytof_20,mcytof_21)
colnames(mcytof) <- c("cell_type", "subject", "value", "year")
```

### Kolmogorov-Smirnoff test comparing each cell freq pop. dist. from 2020 v 2021
```{r}
k.s <- data.frame(matrix(nrow=nrow(cytof_20), ncol=3))
rownames(k.s) <- rownames(cytof_20)
colnames(k.s) <- c('mean_20','mean_21','ks_test')
for (i in 1:nrow(cytof_20)) {
    cell_20 <- cytof_20[i,]
    cell_21 <- cytof_21[i,]
    k.s[i,"ks_test"] <- ks.test(cell_20,cell_21)$p.value
    k.s[i,'mean_20'] <- mean(cell_20)
    k.s[i,'mean_21'] <- mean(cell_21)
}
k.s$mean_20 <- round(k.s$mean_20,3)
k.s$mean_21 <- round(k.s$mean_21,3)
k.s$ks_test <- signif(k.s$ks_test,digits=3)
k.s <- k.s[order(nrow(k.s):1),]
```

### Visualize different distributions
```{r}
ggplot(mcytof, aes(x = value, y = cell_type)) +
  geom_density_ridges(aes(x=value, y=cell_type,
                      group = interaction(year,cell_type),
                      fill=year,alpha=0.5)) +
  theme_ridges() 
```
### Perform BatchQC
```{r}
batchQC(cytof_all, batch=batch,
        report_file="batchqc_report_Cytof_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

### Run ComBat (non-parametric) for cell frequency
Run non-parametric ComBat since data is not normally distributed and is not count data
```{r}
cytof_adjust <- ComBat(dat=cytof_all, batch=batch, par.prior=FALSE)
```

### Run BatchQC on adjusted data
```{r}
batchQC(cytof_adjust, batch=batch,
        report_file="batchqc_report_Cytof_adjust_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

## Batch QC for Olink
```{r}
# Olink merge
olink_all <- cbind(t(olink_baseline_mat_imputed_20),t(olink_baseline_mat_imputed_21))
batch <- c(rep('2020',nrow(olink_baseline_mat_imputed_20)),rep('2021',nrow(olink_baseline_mat_imputed_21)))
```

```{r}
batchQC(olink_all, batch=batch,
        report_file="batchqc_report_Olink_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

### Run ComBat on Olink data
```{r}
olink_adjust<-ComBat(dat=olink_all, batch=batch)
```
### Run BatchQC on adjusted data
```{r}
batchQC(olink_adjust, batch=batch,
        report_file="batchqc_report_Olink_adjust_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

## BatchQC for Ab titers
```{r}
# Olink merge
abtiters_all<-cbind(t(abtiters_baseline_mat_imputed_20),t(abtiters_baseline_mat_imputed_21))
batch<-c(rep('2020',nrow(abtiters_baseline_mat_imputed_20)),rep('2021',nrow(abtiters_baseline_mat_imputed_21)))
```

```{r}
batchQC(abtiters_all, batch=batch,
        report_file="batchqc_report_Abtiters_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```
Evaluation of batchQC output for abtiters shows no/few batch effects, so no batch corrections is needed.

## BatchQC for RNASeq
```{r}
# RNAseq merge
seqtiters_all <- cbind(t(rnaseq_baseline_mat_imputed_20),t(rnaseq_baseline_mat_imputed_21))
batch <- c(rep('2020',nrow(rnaseq_baseline_mat_imputed_20)),rep('2021',nrow(rnaseq_baseline_mat_imputed_21)))
```

```{r}
batchQC(seqtiters_all, batch=batch,
        report_file="batchqc_report_RNASeq_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

### Run ComBat on RNASeq data
RNASeq data has been log-transformed, so can run ComBat (assuming near-normality of distribution)
```{r}
seqtiters_adjust <- ComBat(dat=seqtiters_all, batch=batch)
```

### Run BatchQC on adjusted data
```{r}
batchQC(seqtiters_adjust, batch=batch,
        report_file="batchqc_report_RNASeq_adjust_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```
