---
title: "Impute CMI-PB data using the MICE algorithm"
author: "Joaquin Reyna (jreyna@lji.org)"
date: "2/22/2023"
output: html_document
params:
    year: 2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r, load-libraries, message=FALSE}
library(tidyverse)
library(tibble)
library(UpSetR)
library(mice)
#library(dplyr)
```

## Set environment values
```{r, set-env}
# set year
if (exists("params") == TRUE) {
    year <- as.integer(params$year)
} else {
    year = 2020
}
message(paste("Final year chosen:", year))

# Set path to the harmonized directory
harmony_dir = "../../../results/main/cmi_pb_datasets/processed/all_versus_all/harmonized/"

# Create output directory
outdir <- "../../../results/main/cmi_pb_datasets/processed/all_versus_all/imputation/"
dir.create(outdir, recursive = TRUE, showWarnings = TRUE)
```


## Import Data
### Initialize the datasets
```{r, initialize-the-datasets}
assays <- c("plasma_ab_titer", "pbmc_cell_frequency", "olink", "pbmc_gene_expression")

debug <- FALSE

# Create a list of datasets with an empty nested list for each year
datasets <- list()
```

### Load each dataset
```{r, load-each-dataset}
for (i in 1:length(assays)) {
  assay <- assays[i]
  message(paste('assay:', assay))
  
  if (assay == 'plasma_ab_titer'){
      harmonized_fn <- file.path(harmony_dir, sprintf("%s.mfi_raw.%d.pivoted.tsv", assay, year))
      
  } else {
      harmonized_fn <- file.path(harmony_dir, sprintf("%s.%d.pivoted.tsv", assay, year))
  }

  message(paste('\tfn:', harmonized_fn))
  
  curr_ds <- read.table(harmonized_fn, header=TRUE, sep='\t')
  if (assay == "pbmc_gene_expression" && debug == TRUE) {
      curr_ds = curr_ds[,1:20]
  }
  
  # remove dups (present for some reason sample 113 day -30)
  curr_ds <- unique(curr_ds, subset = c("subject_id", "planned_day_relative_to_boost")) 
  row.names(curr_ds) <- paste(curr_ds[["subject_id"]],
                                curr_ds[["planned_day_relative_to_boost"]],
                                sep='/')
  
  datasets[[assay]] <- curr_ds
}
```

## Investigate the intersection of available data per year
```{r, investigate-intersections}
list_input <- list()
for (assay in assays){
    list_input[[assay]] <- row.names(datasets[[assay]])
}

invisible({
    img_fn = sprintf(file.path(outdir, "upset.%s.svg"), year)
    svg(img_fn)
    upsetplot <- upset(fromList(list_input), order.by = "freq")
    dev.off()
})
upsetplot
```

#### SKIPPED: Heatmap of 2020 data availability (with RNASeq data subset for ease of visualization)
```{r, skipped-1}
# rnaseq.2020_short <- rnaseq.2020[1:36,1:20]
# X_joined_short <- dplyr::full_join(rownames_to_column(log2(as.data.frame(rnaseq.2020_short + 1))),
#                              rownames_to_column(cytof.2020), 
#                              by = "rowname")
# X_joined_short <- dplyr::full_join(X_joined_short,
#                                  rownames_to_column(olink.2020),
#                                  by = "rowname")
# X_joined_short <- dplyr::full_join(X_joined_short,
#                                  rownames_to_column(log2(abtiters.2020 + 1)),
#                                  by = "rowname")
# 
# print(dim(X_joined_short))
# X_mat <- t(as.matrix(dplyr::select(X_joined_short, -rowname)))
# 
# ggplot(reshape2::melt(X_mat)) +
#   geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
#   scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
#   scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = FALSE) +
#   ggtitle("RNA-Seq | CyTOF | OLINK") +
#   xlab("Features") +
#   ylab("Baseline Samples") +
#   cowplot::theme_map() +
#   theme(plot.title = element_text(hjust = .5, size = 14),
#         axis.title = element_text(hjust = .5),
#         axis.title.x = element_text(size = 10),
#         axis.title.y = element_text(size = 10, angle = 90))
```


#### SKIPPED: Remove samples with no RNASeq data
```{r, skipped-2}
# seq_u <- union(rownames(rnaseq.2020),rownames(olink.2020))
# seq_u_c <- union(seq_u,rownames(cytof.2020))
# samples_remove <- setdiff(rownames(abtiters.2020),seq_u_c)
# abtiters.2020_filt <- subset(abtiters.2020,!(rownames(abtiters.2020) %in% samples_remove))
# 
# samples_remove <- setdiff(union(rownames(olink.2020),rownames(cytof.2020)),rownames(rnaseq.2020))
# olink.2020_filt <- subset(olink.2020,!(rownames(olink.2020) %in% samples_remove))
# cytof.2020_filt <- subset(cytof.2020,!(rownames(cytof.2020) %in% samples_remove))
# abtiters.2020_filt <- subset(abtiters.2020_filt,!(rownames(abtiters.2020_filt) %in% samples_remove))
# 
# rnaseq.2020_short = rnaseq.2020[1:36,1:20]
# X_joined <- dplyr::full_join(rownames_to_column(log2(rnaseq.2020_short+1)),
#                              rownames_to_column(cytof.2020_filt),
#                              by = "rowname")
# print(dim(X_joined))
# X_joined <- dplyr::full_join(X_joined,
#                            rownames_to_column(olink.2020_filt),
#                            by = "rowname")
# X_joined <- dplyr::full_join(X_joined,
#                            rownames_to_column(log2(abtiters.2020_filt+1)),
#                            by = "rowname")
# print(dim(X_joined))
#             
# 
# X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))
# 
# ggplot(reshape2::melt(X_mat)) +
#   geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
#   scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
#   scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = FALSE) +
#   ggtitle("RNASeq | CyTOF | OLINK | abtiters") +
#   xlab("Features") +
#   ylab("Baseline Samples") +
#   cowplot::theme_map() +
#   theme(plot.title = element_text(hjust = .5, size = 14),
#         axis.title = element_text(hjust = .5),
#         axis.title.x = element_text(size = 10),
#         axis.title.y = element_text(size = 10, angle = 90))
```

## Impute data using MICE algorithm
#### Filter samples to match RNA-seq
```{r, filter-samples}
# find the set of samples to remove
seq_u <- union(rownames(datasets[["pbmc_gene_expression"]]), rownames(datasets[["olink"]]))
seq_u_c <- union(seq_u, rownames(datasets[["pbmc_cell_frequency"]]))
samples_remove <- setdiff(rownames(datasets[["plasma_ab_titer"]]), seq_u_c)

# filter abtiters
abtiters.2020_filt <- subset(datasets[["plasma_ab_titer"]], !(rownames(datasets[["plasma_ab_titer"]]) %in% samples_remove))

# filter olink and cytof
samples_remove <- setdiff(union(rownames(datasets[["olink"]]),
                                rownames(datasets[["pbmc_cell_frequency"]])),
                                rownames(datasets[["pbmc_gene_expression"]]))
olink.2020_filt <- subset(datasets[["olink"]], !(rownames(datasets[["olink"]]) %in% samples_remove))
cytof.2020_filt <- subset(datasets[["pbmc_cell_frequency"]], !(rownames(datasets[["pbmc_cell_frequency"]]) %in% samples_remove))
abtiters.2020_filt <- subset(abtiters.2020_filt, !(rownames(abtiters.2020_filt) %in% samples_remove))

```
#### Create an input matrix concatenating all datasets

```{r, create-input-matrix}
# create input for merging pbmc_gene_expression
rnaseq_merge = log2(datasets[["pbmc_gene_expression"]] + 1)
rnaseq_merge = rownames_to_column(rnaseq_merge)
rnaseq_merge = subset(rnaseq_merge, select = -c(subject_id, planned_day_relative_to_boost))

# cytof
cytof_merge = rownames_to_column(cytof.2020_filt)
cytof_merge = subset(cytof_merge, select = -c(subject_id, planned_day_relative_to_boost))

# olink
olink_merge = rownames_to_column(olink.2020_filt)
olink_merge = subset(olink_merge, select = -c(subject_id, planned_day_relative_to_boost))

# abtiters
abtiters_merge = rownames_to_column(log2(abtiters.2020_filt + 1))
abtiters_merge = subset(abtiters_merge, select = -c(subject_id, planned_day_relative_to_boost))
```

```{r, create-input-matrix-part2}
# join rnaseq and cytof
X_joined <- dplyr::full_join(rnaseq_merge, cytof_merge, by = "rowname")
print(dim(X_joined))

# add the olink data
X_joined <- dplyr::full_join(X_joined, olink_merge, by = "rowname")
print(dim(X_joined))

# add the ab titer data
X_joined <- dplyr::full_join(X_joined, abtiters_merge, by = "rowname")
print(dim(X_joined))
            
# make the final matrix for imputation by 
# transposing and removing rowname 
X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))
#X_mat <- t(as.matrix(subset(X_joined, select = -rowname)))

print(dim(X_mat))
```

#### Run the imputation
This function will take a long time to run especially given the number of 
imputation datasets generated and the numbe of maximum iterations. 
```{r, run-imputation}
if (debug == TRUE){
    imputed_Data <- mice(X_mat, m=1, maxit = 1, method = 'pmm', seed = 500)
} else {
    imputed_Data <- mice(X_mat, m=4, maxit = 25, method = 'pmm', seed = 500)
}
```

## Save the imputated data
#### Extract the imputed data
```{r, extract-imputations}
X_mat_complete <- complete(imputed_Data, 1)
colnames(X_mat_complete) <- X_joined$rowname

X_mat_complete_df <- as.data.frame(t(X_mat_complete))

split_rownames <- strsplit(row.names(X_mat_complete_df), "/")
X_mat_complete_df["subject_id"] <- sapply(split_rownames, "[", 1)
X_mat_complete_df["planned_day_relative_to_boost"] <- sapply(split_rownames, "[", 2)
```

#### Split the imputed data into dataframes
```{r, split-imputations}
# Split up into datasets:
# rnaseq
rnaseq_cols = colnames(datasets[["pbmc_gene_expression"]])
rnaseq_imputed <- X_mat_complete_df[rnaseq_cols]

# cytof
cytof_cols = colnames(datasets[["pbmc_cell_frequency"]])
cytof_imputed <- X_mat_complete_df[cytof_cols]

# olink
olink_cols = colnames(datasets[["olink"]])
olink_imputed <- X_mat_complete_df[olink_cols]

# abtiters
abtiters_cols = colnames(datasets[["plasma_ab_titer"]])
abtiters_imputed <- X_mat_complete_df[abtiters_cols]
```

#### Visualize imputed data
```{r, visualize-imputations}
X_joined <- dplyr::full_join(rownames_to_column(rnaseq_imputed[,3:20]),
                             rownames_to_column(cytof_imputed[,3:ncol(cytof_imputed)]),
                             by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(olink_imputed[,3:ncol(olink_imputed)]),
                           by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(abtiters_imputed[,3:ncol(abtiters_imputed)]),
                           by = "rowname")

X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))

heatmap <- ggplot(reshape2::melt(X_mat)) +
  geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
  scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = "none") +
  ggtitle("RNASeq | CyTOF | OLINK | abtiters") +
  xlab("Features") +
  ylab("Baseline Samples") +
  cowplot::theme_map() +
  theme(plot.title = element_text(hjust = .5, size = 14),
        axis.title = element_text(hjust = .5),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90))
   
invisible({
    img_fn = sprintf(file.path(outdir, "imputed_heatmap.%s.svg"), year)
    heatmap
    svg(img_fn)
    dev.off()
})
heatmap
```

#### Save to file
```{r, save-to-file}
template_filename = file.path(outdir, '%s.%d.tsv')

rnaseq_filename = sprintf(template_filename, "pbmc_gene_expression", year)
write_tsv(rnaseq_imputed, rnaseq_filename)

cytof_filename = sprintf(template_filename, "pbmc_cell_frequency", year)
write_tsv(cytof_imputed, cytof_filename)

olink_filename = sprintf(template_filename, "olink", year)
write_tsv(olink_imputed, olink_filename)

abtiters_filename = sprintf(template_filename, "plasma_ab_titer", year)
write_tsv(abtiters_imputed, abtiters_filename)
```